### 2019-05-14 msu-workshop-2019.Rmd
add initial files

---
title: "Won’t you be my neighboR?"
subtitle: "An introduction to R for data science in education"
author: "Joshua Rosenberg"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(knitr)
```

# Who am I?

.pull-left[
* Assistant Professor: STEM Education, University of Tennessee, Knoxville
* Dad (one-year-old toddler!)
* Primary areas of interest
+ Data science in education (network analytic methods, experience sampling method, computational grounded theory)
+ Data science education (integrating data science and science education)
]

```{r, out.width = "350px", echo = FALSE, fig.align = "center"}
include_graphics("img/kr-joro.jpg")
```

---

# Agenda

1. Why learn R?
1. Getting started with R Studio
1. Computer programming basics
1. Visualizing data
1. Processing data
1. Modeling data
1. Creating reproducible reports
1. Learning and doing more with R

---

# Why learn R?

* It is capable of carrying out basic and complex statistical analyses
* It is able to work with data small (*n* = 30!) and large (*n* = 100,000) efficiently
* It is a programming language and so is quite flexible
* There is a great, inclusive community of users and developers (and teachers)
* It is cross-platform, open-source, and freely-available

---

# Getting started with R Studio

### To download R Studio:

- Visit this page to download R studio: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)
- Find your operating system (Mac, Windows, or Linux)
- Download the 'latest release' on the page for your operating system and download and install the application

If you do have issues, consider this [page](https://datacarpentry.org/R-ecology-lesson/), and then reach out for help. One good place to start is the [R Studio Community](https://community.rstudio.com/) is a great place to start.

### Check that it worked

Open R Studio.  Find the console window and type in `2 + 2`. If what you can guess is returned (hint: it's what you expect!), then R Studio *and* R both work.

---
## Creating Projects

Before proceeding, we're going to take a few steps to set ourselves to make the analysis easier; namely, through the use of Projects, an R Studio-specific organizational tool.

To create a project, in R Studio, navigate to "File" and then "New Directory".

Then, click "New Project". Choose a directory name for the project that helps you to remember that this is a project that involves data science in education; it can be convenient if the name is typed in `lower-case-letters-separated-by-dashes`, like that. You can also choose the sub-directory. If you are just using this to learn and to test out creating a project, you may consider placing it in your downloads or another temporary directory so that you remember to remove it later.

Even if you do not create a Project, you can always check where your working directory (i.e., where your R is pointing) is by running `getwd()`. To change it manually, run `setwd(desired/file/path/here)`.

---

## Packages

"Packages" are shareable collections of R code that provide functions (i.e., a command to perform a specific task), data and documentation,. Packages increase the functionality of R by improving and expanding on base R (basic R functions).

### Installing and Loading Packages

To download a package, you must call `install.packages()`:

```{r, eval = FALSE}
install.packages("dplyr", repos = "http://cran.us.r-project.org")
```

You can also navigate to the Packages pane, and then click "Install", which will work the same as the line of code above. This is a way to install a package using code or part of the R Studio interface. Usually, writing code is a bit quicker, but using the interface can be very useful and complimentary to use of code.

---

*After* the package is installed, it must be loaded into your R Studio session using `library()`:

```{r}
library(dplyr)
```

We only have to install a package once, but to use it, we have to load it each time we start a new R session.

> a package is a like a book, a library is like a library; you use library() to check a package out of the library
> - Hadley Wickham, Chief Scientist, R Studio

---

### Running Functions from Packages

Once you have loaded the package in your session, you can run the functions that are contained within that package. To find a list of all those functions, you can run this in the R Studio console:

```{r}
help(package = dplyr)
```

The documentation should tell you what the function does, what arguments (i.e., details) needed for it to successfully run, examples, and what the output should look like.

---

If you know the specific function that you want to look up, you can run this in the R Studio console:

```{r, eval = FALSE}
??dplyr::filter
```

---

Once you know what you want to do with the function, you can run it in your code:

```{r}
dat <- # example data frame
  data.frame(stringsAsFactors=FALSE,
             letter = c("A", "A", "A", "B", "B"),
             number = c(1L, 2L, 3L, 4L, 5L))

dat

filter(dat, letter == "A") # using dplyr::filter
```

---

# Computer programming basics

---

# Data structures

- Data Frame

````{r}
data.frame(var1 = c("a", "b", "c"),
           var2 = c(4, 2, 5))
```

- Tibble

```{r}
tibble(var1 = c("a", "b", "c"),
       var2 = c(4, 2, 5))
```

---

# Conditional logic

- `ifelse()`

```{r}
vec <- runif(10) # generates random number between 0 and 1
ifelse(vec >= .5, "greater", "lesser")
```

- `dplyr::case_when()`

```{r}
d1 <- tibble(var1 <- runif(10))
d1 %>% 
  mutate(var2 = case_when(
    vec >= .5 ~ "greater",
    vec < .5 ~ "lesser"
  ))
```

---

# Functions

---

# Iteration

---

# Visualizing data

---

# Visualizing distributions

---

# Comparing many means

---

# Relationships between two variables

---

# Relationships between three variables

---

# Processing data

### Track Two: Welcome to the Tidyverse

The Tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data. Tidy data has a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.

The packages contained in the Tidyverse provide useful functions that augment base R functionality.

You can installing and load the complete Tidyverse with:

```{r, eval = FALSE}
install.packages("tidyverse")
```

```{r, message = FALSE}
library(tidyverse)
```

**For more information on tidy data, check out [Hadley Wickhams's Tidy Data paper](http://vita.had.co.nz/papers/tidy-data.html).**
**For more information on the Tidyverse, check out DataCamp's [Getting Started with the Tidyverse tutorial]( https://www.datacamp.com/community/tutorials/tidyverse-tutorial-r).**

## Loading Data from Various Sources

In this section, we'll load data.

You might be thinking that an Excel file is the first that we would load, but there happens to be a format which you can open and edit in Excel that is even easier to use between Excel and R as well as SPSS and other statistical software, like MPlus, and even other programming languages, like Python. That format is CSV, or a comma-separated-values file.

The CSV file is useful because you can open it with Excel and save Excel files as CSV files. Additionally, and as its name indicates, a CSV file is rows of a spreadsheet with the columns separated by commas, so you can view it in a text editor, like TextEdit for Macintosh, as well. Not surprisingly, Google Sheets easily converts CSV files into a Sheet, and also easily saves Sheets as CSV files.

For these reasons, we start with - and emphasize - reading CSV files.

### Saving a File from the Web

You'll need to copy this URL:

`https://goo.gl/bUeMhV`

Here's what it resolves to (it's a CSV file):

`https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`

This next chunk of code downloads the file to your working directory. Run this to download it so in the next step you can read it into R. As a note: There are ways to read the file directory (from the web) into R. Also, of course, you could do what the next (two) lines of code do manually: Feel free to open the file in your browser and to save it to your computer (you should be able to 'right' or 'control' click the page to save it as a text file with a CSV extension).

```{r, eval = FALSE}
student_responses_url <-
  "https://goo.gl/bUeMhV"

student_responses_file_name <-
  paste0(getwd(), "/data/student-responses-data.csv")

download.file(
  url = student_responses_url,
  destfile = student_responses_file_name)
```

It may take a few seconds to download as it's around 20 MB.

The process above involves many core data science ideas and ideas from programming/coding. We will walk through them step-by-step.

1. The *character string* `"https://goo.gl/wPmujv"` is being saved to an *object* called `student_responses_url`.

```{r}
student_responses_url <-
  "https://goo.gl/bUeMhV"
```

2. We concatenate your working directory file path to the desired file name for the CSV using a *function* called `paste0`. This is stored in another *object* called `student_reponses_file_name`. This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.

```{r}
student_responses_file_name <-
  paste0(getwd(), "/data/student-responses-data.csv")
```

3. The `student_responses_url` *object* is passed to the `url` argument of the *function* called `download.file()` along with `student_responses_file_name`, which is passed to the `destfile` argument.

In short, the `download.file()` function needs to know
- where the file is coming from (which you tell it through the `url`) argument and
- where the file will be saved (which you tell it through the `destfile` argument).

```{r}
download.file(
  url = student_responses_url,
  destfile = student_responses_file_name)
```

Understanding how R is working in these terms can be helpful for troubleshooting and reaching out for help. It also helps you to use functions that you have never used before because you are familiar with how some functions work.

Now, in R Studio, you should see the downloaded file in the Files tab. This should be the case if you created a project with R Studio; if not, it should be whatever your working directory is set to. If the file is there, great. If things are *not* working, consider downloading the file in the manual way and then move it into the directory that the R Project you created it.

### Loading a CSV File

Okay, we're ready to go. The easiest way to read a CSV file is with the function `read_csv()` from the package `readr`, which is contained within the Tidyverse.

Let's load the tidyverse library:

```{r, message = F}
library(tidyverse) # so tidyverse packages can be used for analysis
```

You may have noticed the hash symbol after the code that says `library(tidyverse`)`. It reads `# so tidyverse packages can be used for analysis`. That is a comment and the code after it (but not before it) is not run (the code before it runs just like normal). Comments are useful for showing *why* a line of code does what it does.

After loading the tidyverse packages, we can now load a file. We are going to call the data `student_responses`:

```{r, eval = TRUE}
# readr::write_csv(pisaUSA15::stu_quest, here::here("data", "pisaUSA15", "stu_quest.csv"))
student_responses <-
  read_csv("./data/student-responses-data.csv")
```

Since we loaded the data, we now want to look at it. We can type its name in the function `glimpse()` to print some information on the dataset (this code is not run here).

```{r, eval = FALSE}
glimpse(student_responses)
```

Woah, that's a big data frame (with a lot of variables with confusing names, to boot)!

Great job loading a file and printing it! We are now well on our way to carrying out analysis of our data.

### Loading Excel files

We will now do the same with an Excel file. You might be thinking that you can open the file in Excel and then save it as a CSV. This is generally a good idea. At the same time, sometimes you may need to directly read a file from Excel. Note that, when possible, we recommend the use of CSV files. They work well across platforms and software (i.e., even if you need to load the file with some other software, such as Python).

The package for loading Excel files, `readxl`, is not a part of the tidyverse, so we will have to install it first (remember, we only need to do this once), and then load it using `library(readxl)`. Note that the command to install `readxl` is grayed-out below: The `#` symbol before `install.packages("readxl")` indicates that this line should be treated as a comment and not actually run, like the lines of code that are not grayed-out. It is here just as a reminder that the package needs to be installed if it is not already.

Once we have installed readxl, we have to load it (just like tidyverse):

```{r, eval = FALSE}
install.packages("readxl")
```

```{r}
library(readxl)
```

We can then use the function `read_excel()` in the same way as `read_csv()`, where "path/to/file.xlsx" is where an Excel file you want to load is located (note that this code is not run here):

```{r, eval = FALSE}
my_data <-
  read_excel("path/to/file.xlsx")
```

Of course, were this run, you can replace `my_data` with a name you like. Generally, it's best to use short and easy-to-type names for data as you will be typing and using it a lot.

Note that one easy way to find the path to a file is to use the "Import Dataset" menu. It is in the Environment window of R Studio. Click on that menu bar option, select the option corresponding to the type of file you are trying to load (e.g., "From Excel"), and then click The "Browse" button beside the File/URL field. Once you click on the, R Studio will automatically generate the file path - and the code to read the file, too - for you. You can copy this code or click Import to load the data.

### Loading SAV files

The same factors that apply to reading Excel files apply to reading `SAV` files (from SPSS). NOte that you can also read CSV file directly into SPSS and so because of this and the benefits of using CSVs (they are simple files that work across platforms and software), we recommend using CSVs when possible. First, install the package `haven`, load it, and the use the function `read_sav()`:

```{r, eval = FALSE}
install.packages("haven")
```

```{r, eval = FALSE}
library(haven)
my_data <-
  read_sav("path/to/file.xlsx")
```

### Google Sheets

Finally, it can sometimes be useful to load a file directly from Google Sheets, and this can be done using the Google Sheets package.

```{r, eval = FALSE}
install.packages("googlesheets")
```

```{r, eval = FALSE}
library(googlesheets)
```

When you run the command below, a link to authenticate with your Google account will open in your browser.

```{r, eval = FALSE}
my_sheets <- gs_ls()
```

You can then simply use the `gs_title()` function in conjunction with the `gs_read()` function:

```{r, eval = FALSE}
df <- gs_title('title')
df <- gs_read(df)
```

### Saving Files

Using our data frame `student_responses`, we can save it as a CSV (for example) with the following function. The first argument, `student_reponses`, is the name of the object that you want to save. The second argument, `student-responses.csv`, what you want to call the saved dataset.

```{r, eval = FALSE}
write_csv(student_responses, "student-responses.csv")
```

That will save a CSV file entitled `student-responses.csv` in the working directory. If you want to save it to another directory, simply add the file path to the file, i.e. `path/to/student-responses.csv`. To save a file for SPSS, load the haven package and use `write_sav()`. There is not a function to save an Excel file, but you can save as a CSV and directly load it in Excel.

### Conclusion

We will detail the functions used to read every file in a folder (or, to write files to a folder).

**For more information on installing R and R Studio, check out DataCamp's [R Data Import Tutorial](https://www.datacamp.com/community/tutorials/installing-R-windows-mac-ubuntu).**

## Processing Data

Now that we have loaded `student_responses` into an object, we can process it. This section highlights some common data processing functions.

We're also going to introduce a powerful, unusual *operator* in R, the pipe. The pipe is this symbol: `%>%`. It lets you *compose* functions. It does this by passing the output of one function to the next. A handy shortcut for writing out `%>%` is Command + Shift + M.

Here's an example. Let's say that we want to select a few variables from the `student_responses` dataset and save those variables into a new object, `student_mot_vars`. Here's how we would do that using `dplyr::select()`.

```{r}
student_mot_vars <- # save object student_mot_vars by...
  student_responses %>% # using dataframe student_responses
  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
```

Note that we saved the output from the `select()` function to `student_mot_vars` but we could also save it back to `student_responses`, which would simply overwrite the original data frame (the following code is not run here):

```{r, eval = FALSE}
student_responses <- # save object student_responses by...
  student_responses %>% # using dataframe student_responses
  select(student_responses, SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
```

We can also rename the variables at the same time we select them. I put these on separate lines so I could add the comment, but you could do this all in the same line, too. It does not make a difference in terms of how `select()` will work.

```{r}
student_mot_vars <- # save object student_mot_vars by...
  student_responses %>% # using dataframe student_responses
  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
         student_instrumental_motivation = INSTSCIE # selecting variable INSTSCIE and renaming to student_instrumental_motivation
  )
```

[will add more on creating new variables, filtering grouping and summarizing, and joining data sets]

## Communicating / sharing results

R Markdown is a highly convenient way to communicate and share results. Navigate to "New File" and then "R Markdown". [add]

Then, click "Knit to PDF", "Knit to HTML", or "Knit to Word".

## Other foundational notes

### Configuring R Studio

There are a number of changes you *can* (but do not need to) make to configure R Studio. If you navigate to the Preferences menu in R Studio, you'll see a number of options you can change, from the appearance of the application to which windows appear where.

One important consideration is whether to save your workspace when you close R Studio. By default, R Studio saves all of the objects in your environment. This means that any data that you have loaded--or new data or objects that you have created, such as by merging two data sets together or creating a plot--will, by default, still exist when you open R Studio next. In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code. This means that when you share your code, or re-run it from the start, it may not work. An easy way to change this is to tell R Studio to start from scratch (in terms of your workspace) each time you open it. You can do that by changing the dropdown menu pointed out in the image below to "Never".

![optional caption text](images/save-workspace-reminder.jpg)

While this may seem like a dramatic step - never saving your workspace - it is the foundation for doing reproducible work and research using R Studio (and R). It also represents one of the biggest shifts from using software like Excel or SPSS, where most steps are not documented in code. This involves a shift from thinking that your most permanent and important part of an analysis is your data to thinking of the most important part as being the code: with the code, you can keep your data in its original form, process it, and then save a processed file, through running code. This also means that when you have to make a change to this code, you can re-run the entire analysis easily.

---

# Processing data (part 1 of 4)

* `select()`
* `rename()`
* `filter()`
* `mutate()`

---

# Processing data (part 2 of 4)

* `group_by()`
* `count()`
* `summarize()`

---

# Processing data (part 3 of 4)

* `left_join()`
* `right_join()`
* `semi_join()`
* `anti_join()`

---

# Processing data (part 4 of 4)

* `gather()`
* `spread()`

---

# Modeling data

---

# Linear models

---

# Multi-level models

---

# Latent variable models

---

# Creating reproducible reports

---

# R Markdown

---

# APA Tables

---

# Papaja

---

# Redoc

---

# Learning and doing more with R

---

# Following up

### 2019-05-15 msu-workshop-2019.Rmd
update post conversation with Emily

@@
 library(knitr)
+library(tidyverse)
 ```
@@
 ```
+
+---
+
+class: inverse, center, middle
+
+# Agenda
 
@@
 
+class: inverse, center, middle
+
 # Getting started with R Studio
+
+---
 
@@
 
+# Using packages
+
 *After* the package is installed, it must be loaded into your R Studio session using `library()`:
@@
 
+class: inverse, center, middle
+
 # Computer programming basics
+
+---
+
+# Objects and variables
+
+```{r}
+my_object <- 1
+my_other_object <- c(1:10)
+```
+
+Reading in a CSV and assigning it to an object
+
+```{r}
+my_df <- read_csv("example-data.csv")
+my_df
+```
 
@@
 ```{r}
+library(tidyverse)
 tibble(var1 = c("a", "b", "c"),
@@
 
+# Data structures
+
+```{r}
+vec1 = c("a", "b", "c")
+vec1 
+```
+
+```{r}
+vec2 = c(4, 2, 5)
+vec2
+```
+
+```{r}
+library(tibble)
+tibble(var1 = vec1,
+       var2 = vec2)
+```
+
 # Conditional logic
@@
 ```{r}
-vec <- runif(10) # generates random number between 0 and 1
-ifelse(vec >= .5, "greater", "lesser")
+vec1 <- runif(10) # generates random number between 0 and 1
+d1 <- tibble(var1 = vec1)
+
+d1 %>% 
+  mutate(var2 = ifelse(vec1 >= .5, "greater", "lesser"))
 ```
@@
 ```{r}
-d1 <- tibble(var1 <- runif(10))
 d1 %>% 
   mutate(var2 = case_when(
-    vec >= .5 ~ "greater",
-    vec < .5 ~ "lesser"
+    var1 >= .75 ~ "a lot",
+    var1 >= .5 ~ "greater",
+    var1 < .5 ~ "lesser",
+    var1 <= .25 ~ "very little" 
   ))
@@
 
+```{r}
+mean(c(1, 2, 3))
+```
+
+```{r}
+mean(c(1, 2, 3, NA))
+```
+
+```{r, eval = FALSE}
+?mean
+```
+
+```{r}
+mean(c(1, 2, 3, NA), na.rm = TRUE)
+```
+
+---
+
+# Functions
+
+```{r, eval = FALSE}
+library(tidyverse)
+read_directory <- function(directory) {
+  map_df(list.files(directory), read_csv)
+}
+
+d <- read_directory("data")
+```
+
 ---
@@
 
+```{r}
+random_numbers <- rnorm(n = 10, mean = 0)
+random_numbers
+mean(random_numbers)
+sd(random_numbers)
+```
+
+```{r}
+library(tidyverse)
+1:10 %>%
+  map(~ rnorm(n = 10, mean = .x)) %>%
+  map_dbl(mean)
+```
+
 ---
 
+class: inverse, center, middle
 # Visualizing data
+
+---
 
@@
 ---
+class: inverse, center, middle
+# Processing data
 
-# Processing data
+---
 
@@
 **For more information on tidy data, check out [Hadley Wickhams's Tidy Data paper](http://vita.had.co.nz/papers/tidy-data.html).**
-**For more information on the Tidyverse, check out DataCamp's [Getting Started with the Tidyverse tutorial]( https://www.datacamp.com/community/tutorials/tidyverse-tutorial-r).**
 
@@
 
-**For more information on installing R and R Studio, check out DataCamp's [R Data Import Tutorial](https://www.datacamp.com/community/tutorials/installing-R-windows-mac-ubuntu).**
-
 ## Processing Data
@@
 
+class: inverse, center, middle
+
 # Modeling data
@@
 
+class: inverse, center, middle
 # Creating reproducible reports

### 2019-05-16 msu-workshop-2019.Rmd
update

@@
 1. Getting started with R Studio
-1. Computer programming basics
 1. Visualizing data
 1. Processing data
-1. Modeling data
-1. Creating reproducible reports
 1. Learning and doing more with R
+1. Modeling data (time permitting)
+1. Computer programming basics (time permitting)
+1. Creating reproducible reports (time permitting)
 
@@
 * It is capable of carrying out basic and complex statistical analyses
-* It is able to work with data small (*n* = 30!) and large (*n* = 100,000) efficiently
+* It is able to work with data small (*n* = 30!) and large (*n* = 100,000+) efficiently
 * It is a programming language and so is quite flexible
@@
 
-### To download R Studio:
+# To download R Studio:
 
@@
 ---
-## Creating Projects
+# Creating Projects
 
@@
 
-To create a project, in R Studio, navigate to "File" and then "New Directory".
-
-Then, click "New Project".
-Choose a directory name for the project that helps you to remember that this is a project that involves data science in education; it can be convenient if the name is typed in `lower-case-letters-separated-by-dashes`, like that.
-You can also choose the sub-directory.
-If you are just using this to learn and to test out creating a project, you may consider placing it in your downloads or another temporary directory so that you remember to remove it later.
-
-Even if you do not create a Project, you can always check where your working directory (i.e., where your R is pointing) is by running `getwd()`.
-To change it manually, run `setwd(desired/file/path/here)`.
+To create a project, in R Studio, navigate to "File" and then "New Directory". Then, click "New Project".
 
@@
 
-## Packages
+# Packages
 
@@
 
-### Running Functions from Packages
+# Running Functions from Packages
 
@@
 
-```{r}
+```{r, eval = FALSE}
 help(package = dplyr)
@@
 
+# Vignettes
+
+```{r, eval = FALSE}
+vignette(package = "dplyr")
+vignette("dplyr", package = "dplyr")
+```
+
+---
+
+# Using a specific function
+
 If you know the specific function that you want to look up, you can run this in the R Studio console:
@@
 ```
-
----
 
@@
 ```
+
+---
+
+# Configuring R Studio
+
+One important consideration is whether to save your workspace when you close R Studio.
+By default, R Studio saves all of the objects in your environment.
+This means that any data that you have loaded--or new data or objects that you have created, such as by merging two data sets together or creating a plot--will, by default, still exist when you open R Studio next.
+In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code.
+This means that when you share your code, or re-run it from the start, it may not work.
+An easy way to change this is to tell R Studio to start from scratch (in terms of your workspace) each time you open it.
+You can do that by changing the dropdown menu pointed out in the image below to "Never".
+
+![optional caption text](images/save-workspace-reminder.jpg)
+---
+
+class: inverse, center, middle
+# Processing data
+
+---
+
+# The Tidyverse
+
+The Tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data.
+Tidy data has a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.
+
+The packages contained in the Tidyverse provide useful functions that augment base R functionality.
+
+You can installing and load the complete tidyverse with:
+
+```{r, eval = FALSE}
+install.packages("tidyverse")
+```
+
+```{r, message = FALSE}
+library(tidyverse)
+```
+
+**For more information on tidy data, check out [Hadley Wickhams's Tidy Data paper](http://vita.had.co.nz/papers/tidy-data.html).**
+---
+
+# Loading Data from Various Sources
+
+### Saving a File from the Web
+
+You'll need to copy this URL:
+
+`https://goo.gl/bUeMhV`
+
+Here's what it resolves to (it's a CSV file):
+
+`https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`
+
+```{r, eval = FALSE}
+student_responses_url <-
+  "https://goo.gl/bUeMhV"
+
+student_responses_file_name <-
+  paste0(getwd(), "./data/student-responses-data.csv")
+
+download.file(
+  url = student_responses_url,
+  destfile = student_responses_file_name)
+```
+
+It may take a few seconds to download as it's around 20 MB.
+
+The process above involves many core data science ideas and ideas from programming/coding. We will walk through them step-by-step.
+
+---
+
+# Step 1
+
+1. The *character string* `"https://goo.gl/wPmujv"` is being saved to an *object* called `student_responses_url`.
+
+```{r}
+student_responses_url <-
+  "https://goo.gl/bUeMhV"
+```
+
+---
+
+# Step 2
+
+2.
+We concatenate your working directory file path to the desired file name for the CSV using a *function* called `paste0`.
+This is stored in another *object* called `student_reponses_file_name`.
+This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.
+
+```{r}
+student_responses_file_name <-
+  paste0(getwd(), "./data/student-responses-data.csv")
+```
+
+---
+
+# Step 3
+
+3. The `student_responses_url` *object* is passed to the `url` argument of the *function* called `download.file()` along with `student_responses_file_name`, which is passed to the `destfile` argument.
+
+In short, the `download.file()` function needs to know
+- where the file is coming from (which you tell it through the `url`) argument and
+- where the file will be saved (which you tell it through the `destfile` argument).
+
+```{r, eval = FALSE}
+download.file(
+  url = student_responses_url,
+  destfile = student_responses_file_name)
+```
+
+Understanding how R is working in these terms can be helpful for troubleshooting and reaching out for help.
+It also helps you to use functions that you have never used before because you are familiar with how some functions work.
+
+---
+
+# Loading a CSV File
+
+Okay, we're ready to go. The easiest way to read a CSV file is with the function `read_csv()` from the package `readr`, which is contained within the Tidyverse.
+
+---
+
+# Loading the file
+
+Let's load the tidyverse library:
+
+```{r, message = F}
+library(tidyverse) # so tidyverse packages can be used for analysis
+```
+
+After loading the tidyverse packages, we can now load a file. We are going to call the data `student_responses`:
+
+```{r, eval = TRUE}
+student_responses <-
+  read_csv("./data/student-responses-data.csv")
+```
+
+Since we loaded the data, we now want to look at it. We can type its name in the function `glimpse()` to print some information on the dataset (this code is not run here).
+
+```{r, eval = FALSE}
+glimpse(student_responses)
+```
+
+---
+
+# Loading Excel files
+
+```{r, eval = FALSE}
+install.packages("readxl")
+```
+
+```{r}
+library(readxl)
+```
+
+We can then use the function `read_excel()` in the same way as `read_csv()`, where "path/to/file.xlsx" is where an Excel file you want to load is located (note that this code is not run here):
+
+```{r, eval = FALSE}
+my_data <-
+  read_excel("path/to/file.xlsx")
+```
+
+---
+
+# Loading SAV files
+
+```{r, eval = FALSE}
+install.packages("haven")
+```
+
+```{r, eval = FALSE}
+library(haven)
+my_data <-
+  read_sav("path/to/file.xlsx")
+```
+
+---
+
+# Saving Files
+
+Using our data frame `student_responses`, we can save it as a CSV (for example) with the following function.
+The first argument, `student_reponses`, is the name of the object that you want to save.
+The second argument, `student-responses.csv`, what you want to call the saved dataset.
+
+```{r, eval = FALSE}
+write_csv(student_responses, "student-responses.csv")
+```
+
+That will save a CSV file entitled `student-responses.csv` in the working directory.
+If you want to save it to another directory, simply add the file path to the file, i.e.
+`path/to/student-responses.csv`.
+To save a file for SPSS, load the haven package and use `write_sav()`.
+There is not a function to save an Excel file, but you can save as a CSV and directly load it in Excel.
+
+---
+class: inverse, center, middle
+# Processing Data
+---
+
+# The pipe: %>%
+
+We're also going to introduce a powerful, unusual *operator* in R, the pipe.
+The pipe is this symbol: `%>%`.
+It lets you *compose* functions.
+It does this by passing the output of one function to the next.
+A handy shortcut for writing out `%>%` is Command + Shift + M.
+
+Here's an example.
+Let's say that we want to select a few variables from the `student_responses` dataset and save those variables into a new object, `student_mot_vars`.
+Here's how we would do that using `dplyr::select()`.
+
+```{r}
+student_mot_vars <- # save object student_mot_vars by...
+  student_responses %>% # using dataframe student_responses
+  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
+```
+
+---
+
+# Saving the results
+
+Note that we saved the output from the `select()` function to `student_mot_vars` but we could also save it back to `student_responses`, which would simply overwrite the original data frame (the following code is not run here):
+
+```{r, eval = FALSE}
+student_responses <- # save object student_responses by...
+  student_responses %>% # using dataframe student_responses
+  select(student_responses, SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
+```
+
+---
+
+# Renaming
+
+We can also rename the variables at the same time we select them.
+I put these on separate lines so I could add the comment, but you could do this all in the same line, too.
+It does not make a difference in terms of how `select()` will work.
+
+```{r}
+student_mot_vars <- # save object student_mot_vars by...
+  student_responses %>% # using dataframe student_responses
+  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
+         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
+         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
+         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
+         student_instrumental_motivation = INSTSCIE # selecting variable INSTSCIE and renaming to student_instrumental_motivation
+  )
+```
+
+---
+
+# Processing data (part 1 of 4)
+
+* `select()`
+* `rename()`
+* `filter()`
+* `mutate()`
+
+---
+
+# Processing data (part 2 of 4)
+
+* `group_by()`
+* `count()`
+* `summarize()`
+
+---
+
+# Processing data (part 3 of 4)
+
+* `left_join()`
+* `right_join()`
+* `semi_join()`
+* `anti_join()`
+
+---
+
+# Processing data (part 4 of 4)
+
+* `gather()`
+* `spread()`
+
+---
+
+class: inverse, center, middle
+
+# Modeling data
+
+---
+
+# Linear models
+
+---
+
+# Multi-level models
+
+---
+
+# Latent variable models
+
+---
+
+class: inverse, center, middle
+# Creating reproducible reports
+
+---
+
+# R Markdown
+
+---
+
+# APA Tables
+
+---
+
+# Papaja
+
+---
+
+# Redoc
 
@@
 
-class: inverse, center, middle
-# Visualizing data
-
----
-
----
-
-# Visualizing distributions
-
----
-
-# Comparing many means
-
----
-
-# Relationships between two variables
-
----
-
-# Relationships between three variables
-
----
-class: inverse, center, middle
-# Processing data
-
----
-
-### Track Two: Welcome to the Tidyverse
-
-The Tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data.
-Tidy data has a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.
-
-The packages contained in the Tidyverse provide useful functions that augment base R functionality.
-
-You can installing and load the complete Tidyverse with:
-
-```{r, eval = FALSE}
-install.packages("tidyverse")
-```
-
-```{r, message = FALSE}
-library(tidyverse)
-```
-
-**For more information on tidy data, check out [Hadley Wickhams's Tidy Data paper](http://vita.had.co.nz/papers/tidy-data.html).**
-
-## Loading Data from Various Sources
-
-In this section, we'll load data.
-
-You might be thinking that an Excel file is the first that we would load, but there happens to be a format which you can open and edit in Excel that is even easier to use between Excel and R as well as SPSS and other statistical software, like MPlus, and even other programming languages, like Python.
-That format is CSV, or a comma-separated-values file.
-
-The CSV file is useful because you can open it with Excel and save Excel files as CSV files.
-Additionally, and as its name indicates, a CSV file is rows of a spreadsheet with the columns separated by commas, so you can view it in a text editor, like TextEdit for Macintosh, as well.
-Not surprisingly, Google Sheets easily converts CSV files into a Sheet, and also easily saves Sheets as CSV files.
-
-For these reasons, we start with - and emphasize - reading CSV files.
-
-### Saving a File from the Web
-
-You'll need to copy this URL:
-
-`https://goo.gl/bUeMhV`
-
-Here's what it resolves to (it's a CSV file):
-
-`https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`
-
-This next chunk of code downloads the file to your working directory.
-Run this to download it so in the next step you can read it into R.
-As a note: There are ways to read the file directory (from the web) into R.
-Also, of course, you could do what the next (two) lines of code do manually: Feel free to open the file in your browser and to save it to your computer (you should be able to 'right' or 'control' click the page to save it as a text file with a CSV extension).
-
-```{r, eval = FALSE}
-student_responses_url <-
-  "https://goo.gl/bUeMhV"
-
-student_responses_file_name <-
-  paste0(getwd(), "/data/student-responses-data.csv")
-
-download.file(
-  url = student_responses_url,
-  destfile = student_responses_file_name)
-```
-
-It may take a few seconds to download as it's around 20 MB.
-
-The process above involves many core data science ideas and ideas from programming/coding. We will walk through them step-by-step.
-
-1. The *character string* `"https://goo.gl/wPmujv"` is being saved to an *object* called `student_responses_url`.
-
-```{r}
-student_responses_url <-
-  "https://goo.gl/bUeMhV"
-```
-
-2.
-We concatenate your working directory file path to the desired file name for the CSV using a *function* called `paste0`.
-This is stored in another *object* called `student_reponses_file_name`.
-This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.
-
-```{r}
-student_responses_file_name <-
-  paste0(getwd(), "/data/student-responses-data.csv")
-```
-
-3. The `student_responses_url` *object* is passed to the `url` argument of the *function* called `download.file()` along with `student_responses_file_name`, which is passed to the `destfile` argument.
-
-In short, the `download.file()` function needs to know
-- where the file is coming from (which you tell it through the `url`) argument and
-- where the file will be saved (which you tell it through the `destfile` argument).
-
-```{r}
-download.file(
-  url = student_responses_url,
-  destfile = student_responses_file_name)
-```
-
-Understanding how R is working in these terms can be helpful for troubleshooting and reaching out for help.
-It also helps you to use functions that you have never used before because you are familiar with how some functions work.
-
-Now, in R Studio, you should see the downloaded file in the Files tab.
-This should be the case if you created a project with R Studio; if not, it should be whatever your working directory is set to.
-If the file is there, great.
-If things are *not* working, consider downloading the file in the manual way and then move it into the directory that the R Project you created it.
-
-### Loading a CSV File
-
-Okay, we're ready to go. The easiest way to read a CSV file is with the function `read_csv()` from the package `readr`, which is contained within the Tidyverse.
-
-Let's load the tidyverse library:
-
-```{r, message = F}
-library(tidyverse) # so tidyverse packages can be used for analysis
-```
-
-You may have noticed the hash symbol after the code that says `library(tidyverse`)`.
-It reads `# so tidyverse packages can be used for analysis`.
-That is a comment and the code after it (but not before it) is not run (the code before it runs just like normal).
-Comments are useful for showing *why* a line of code does what it does.
-
-After loading the tidyverse packages, we can now load a file. We are going to call the data `student_responses`:
-
-```{r, eval = TRUE}
-# readr::write_csv(pisaUSA15::stu_quest, here::here("data", "pisaUSA15", "stu_quest.csv"))
-student_responses <-
-  read_csv("./data/student-responses-data.csv")
-```
-
-Since we loaded the data, we now want to look at it. We can type its name in the function `glimpse()` to print some information on the dataset (this code is not run here).
-
-```{r, eval = FALSE}
-glimpse(student_responses)
-```
-
-Woah, that's a big data frame (with a lot of variables with confusing names, to boot)!
-
-Great job loading a file and printing it! We are now well on our way to carrying out analysis of our data.
-
-### Loading Excel files
-
-We will now do the same with an Excel file.
-You might be thinking that you can open the file in Excel and then save it as a CSV.
-This is generally a good idea.
-At the same time, sometimes you may need to directly read a file from Excel.
-Note that, when possible, we recommend the use of CSV files.
-They work well across platforms and software (i.e., even if you need to load the file with some other software, such as Python).
-
-The package for loading Excel files, `readxl`, is not a part of the tidyverse, so we will have to install it first (remember, we only need to do this once), and then load it using `library(readxl)`.
-Note that the command to install `readxl` is grayed-out below: The `#` symbol before `install.packages("readxl")` indicates that this line should be treated as a comment and not actually run, like the lines of code that are not grayed-out.
-It is here just as a reminder that the package needs to be installed if it is not already.
-
-Once we have installed readxl, we have to load it (just like tidyverse):
-
-```{r, eval = FALSE}
-install.packages("readxl")
-```
-
-```{r}
-library(readxl)
-```
-
-We can then use the function `read_excel()` in the same way as `read_csv()`, where "path/to/file.xlsx" is where an Excel file you want to load is located (note that this code is not run here):
-
-```{r, eval = FALSE}
-my_data <-
-  read_excel("path/to/file.xlsx")
-```
-
-Of course, were this run, you can replace `my_data` with a name you like. Generally, it's best to use short and easy-to-type names for data as you will be typing and using it a lot.
-
-Note that one easy way to find the path to a file is to use the "Import Dataset" menu.
-It is in the Environment window of R Studio.
-Click on that menu bar option, select the option corresponding to the type of file you are trying to load (e.g., "From Excel"), and then click The "Browse" button beside the File/URL field.
-Once you click on the, R Studio will automatically generate the file path - and the code to read the file, too - for you.
-You can copy this code or click Import to load the data.
-
-### Loading SAV files
-
-The same factors that apply to reading Excel files apply to reading `SAV` files (from SPSS).
-NOte that you can also read CSV file directly into SPSS and so because of this and the benefits of using CSVs (they are simple files that work across platforms and software), we recommend using CSVs when possible.
-First, install the package `haven`, load it, and the use the function `read_sav()`:
-
-```{r, eval = FALSE}
-install.packages("haven")
-```
-
-```{r, eval = FALSE}
-library(haven)
-my_data <-
-  read_sav("path/to/file.xlsx")
-```
-
-### Google Sheets
-
-Finally, it can sometimes be useful to load a file directly from Google Sheets, and this can be done using the Google Sheets package.
-
-```{r, eval = FALSE}
-install.packages("googlesheets")
-```
-
-```{r, eval = FALSE}
-library(googlesheets)
-```
-
-When you run the command below, a link to authenticate with your Google account will open in your browser.
-
-```{r, eval = FALSE}
-my_sheets <- gs_ls()
-```
-
-You can then simply use the `gs_title()` function in conjunction with the `gs_read()` function:
-
-```{r, eval = FALSE}
-df <- gs_title('title')
-df <- gs_read(df)
-```
-
-### Saving Files
-
-Using our data frame `student_responses`, we can save it as a CSV (for example) with the following function.
-The first argument, `student_reponses`, is the name of the object that you want to save.
-The second argument, `student-responses.csv`, what you want to call the saved dataset.
-
-```{r, eval = FALSE}
-write_csv(student_responses, "student-responses.csv")
-```
-
-That will save a CSV file entitled `student-responses.csv` in the working directory.
-If you want to save it to another directory, simply add the file path to the file, i.e.
-`path/to/student-responses.csv`.
-To save a file for SPSS, load the haven package and use `write_sav()`.
-There is not a function to save an Excel file, but you can save as a CSV and directly load it in Excel.
-
-### Conclusion
-
-We will detail the functions used to read every file in a folder (or, to write files to a folder).
-
-## Processing Data
-
-Now that we have loaded `student_responses` into an object, we can process it. This section highlights some common data processing functions.
-
-We're also going to introduce a powerful, unusual *operator* in R, the pipe.
-The pipe is this symbol: `%>%`.
-It lets you *compose* functions.
-It does this by passing the output of one function to the next.
-A handy shortcut for writing out `%>%` is Command + Shift + M.
-
-Here's an example.
-Let's say that we want to select a few variables from the `student_responses` dataset and save those variables into a new object, `student_mot_vars`.
-Here's how we would do that using `dplyr::select()`.
-
-```{r}
-student_mot_vars <- # save object student_mot_vars by...
-  student_responses %>% # using dataframe student_responses
-  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
-```
-
-Note that we saved the output from the `select()` function to `student_mot_vars` but we could also save it back to `student_responses`, which would simply overwrite the original data frame (the following code is not run here):
-
-```{r, eval = FALSE}
-student_responses <- # save object student_responses by...
-  student_responses %>% # using dataframe student_responses
-  select(student_responses, SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
-```
-
-We can also rename the variables at the same time we select them.
-I put these on separate lines so I could add the comment, but you could do this all in the same line, too.
-It does not make a difference in terms of how `select()` will work.
-
-```{r}
-student_mot_vars <- # save object student_mot_vars by...
-  student_responses %>% # using dataframe student_responses
-  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
-         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
-         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
-         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
-         student_instrumental_motivation = INSTSCIE # selecting variable INSTSCIE and renaming to student_instrumental_motivation
-  )
-```
-
-[will add more on creating new variables, filtering grouping and summarizing, and joining data sets]
-
-## Communicating / sharing results
-
-R Markdown is a highly convenient way to communicate and share results. Navigate to "New File" and then "R Markdown". [add]
-
-Then, click "Knit to PDF", "Knit to HTML", or "Knit to Word".
-
-## Other foundational notes
-
-### Configuring R Studio
-
-There are a number of changes you *can* (but do not need to) make to configure R Studio.
-If you navigate to the Preferences menu in R Studio, you'll see a number of options you can change, from the appearance of the application to which windows appear where.
-
-One important consideration is whether to save your workspace when you close R Studio.
-By default, R Studio saves all of the objects in your environment.
-This means that any data that you have loaded--or new data or objects that you have created, such as by merging two data sets together or creating a plot--will, by default, still exist when you open R Studio next.
-In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code.
-This means that when you share your code, or re-run it from the start, it may not work.
-An easy way to change this is to tell R Studio to start from scratch (in terms of your workspace) each time you open it.
-You can do that by changing the dropdown menu pointed out in the image below to "Never".
-
-![optional caption text](images/save-workspace-reminder.jpg)
-
-While this may seem like a dramatic step - never saving your workspace - it is the foundation for doing reproducible work and research using R Studio (and R).
-It also represents one of the biggest shifts from using software like Excel or SPSS, where most steps are not documented in code.
-This involves a shift from thinking that your most permanent and important part of an analysis is your data to thinking of the most important part as being the code: with the code, you can keep your data in its original form, process it, and then save a processed file, through running code.
-This also means that when you have to make a change to this code, you can re-run the entire analysis easily.
-
----
-
-# Processing data (part 1 of 4)
-
-* `select()`
-* `rename()`
-* `filter()`
-* `mutate()`
-
----
-
-# Processing data (part 2 of 4)
-
-* `group_by()`
-* `count()`
-* `summarize()`
-
----
-
-# Processing data (part 3 of 4)
-
-* `left_join()`
-* `right_join()`
-* `semi_join()`
-* `anti_join()`
-
----
-
-# Processing data (part 4 of 4)
-
-* `gather()`
-* `spread()`
-
----
-
-class: inverse, center, middle
-
-# Modeling data
-
----
-
-# Linear models
-
----
-
-# Multi-level models
-
----
-
-# Latent variable models
-
----
-
-class: inverse, center, middle
-# Creating reproducible reports
-
----
-
-# R Markdown
-
----
-
-# APA Tables
-
----
-
-# Papaja
-
----
-
-# Redoc
-
----
-
 # Learning and doing more with R

### 2019-05-16 demo-doc.Rmd
add demo doc

---
title: "Workshop Demo Doc"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
gss_cat
```

# Select

```{r}
gss_cat %>% 
  select(year, marital, tvhours)
```

Try it out!

Select three *different* variables

```{r}
gss_cat %>% 
  select()
```

# Filter

```{r}
gss_cat %>% 
  filter(age > 30)
```

Try it out!

Filter to include only those individuals who watch a lot of TV - more than 10 hours per *day*!

```{r}
gss_cat %>%

```

# Mutate

```{r}

### 2019-05-16 demo-doc.Rmd
update

@@
 
+# Demo doc (part 1): Try it out!
+
+Let's first load the tidyverse suite of packages.
+
+If you haven't already, make sure to install the tidyverse, first.
+
+*You can skip this code "chunk" if you already have the tidyverse installed; you only need to do this once*
+
+```{r}
+install.packages("tidyverse")
+```
+
+Be sure to run this next line to load the tidyverse. You do this each time you open R.
+
 ```{r}
@@
 ```
+
+Here is the data set we will use:
 
@@
 
+Let's select only a few variables.
+
 ```{r}
@@
 
-Try it out!
+**Try it out!**
 
-Select three *different* variables
+Now, select three *different* variables, and run the code chunk to view the result.
 
@@
 
-Try it out!
+**Try it out!**
 
 Filter to include only those individuals who watch a lot of TV - more than 10 hours per *day*!
+
+Be sure to run the result to see whether it did what you think it should do!
 
@@
 
+Let's create a new variable for those 60 years of age or older.
+
 ```{r}
+gss_cat %>% 
+  mutate(potential_senior = if_else(age >= 60, TRUE, FALSE))
+```

### 2019-05-16 msu-workshop-2019.Rmd
update

@@
 
-# Who am I?
+# #whoami1
 
 .pull-left[
-* Assistant Professor: STEM Education, University of Tennessee, Knoxville
+* Dr. Joshua Rosenberg
+* Assistant Professor, STEM Education, University of Tennessee, Knoxville
 * Dad (one-year-old toddler!)
 * Primary areas of interest
-+ Data science in education (network analytic methods, experience sampling method, computational grounded theory)
-+ Data science education (integrating data science and science education)
+  * Data science in education
+  * Data science education
 ]
@@
 include_graphics("img/kr-joro.jpg")
+```
+
+---
+
+# #whoami2
+
+.pull-left[
+* Dr. Emily Bovee
+* Director of Educational Development and Assessment: Marquette University School of Dentistry
+* Cat mom
+* Primary areas of interest
+  * Understanding student motivation in higher education and professional education
+  * Data science education 
+]
+
+```{r, out.width = "350px", echo = FALSE, fig.align = "center"}
+include_graphics("img/ebv.jpg")
 ```
@@
 1. Getting started with R Studio
+1. Computer science foundations
+1. Processing data
+1. Loading data
 1. Visualizing data
-1. Processing data
+1. Modeling data 
 1. Learning and doing more with R
-1. Modeling data (time permitting)
-1. Computer programming basics (time permitting)
-1. Creating reproducible reports (time permitting)
+1. Questions and sundry tools (time permitting)
 
@@
 * It is cross-platform, open-source, and freely-available
-
+* It can h
 ---
@@
 
-# To download R Studio:
+# Installations
 
-- Visit this page to download R studio: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)
+To download R:
+- Visit this page to download R: https://cran.r-project.org/
 - Find your operating system (Mac, Windows, or Linux)
@@
 
-If you do have issues, consider this [page](https://datacarpentry.org/R-ecology-lesson/), and then reach out for help.
-One good place to start is the [R Studio Community](https://community.rstudio.com/) is a great place to start.
+To download R Studio:
+- Visit this page to download R studio: https://www.rstudio.com/products/rstudio/download/
+- Find your operating system (Mac, Windows, or Linux)
+- Download the 'latest release' on the page for your operating system and download and install the application
 
-### Check that it worked
+## Check that it worked
 
@@
 ---
+
 # Creating Projects
@@
 To create a project, in R Studio, navigate to "File" and then "New Directory". Then, click "New Project".
+
+Names should be short and easy-to-remember but also informative.
 
@@
 ```{r, eval = FALSE}
-install.packages("dplyr", repos = "http://cran.us.r-project.org")
+install.packages("dplyr")
 ```
@@
 This is a way to install a package using code or part of the R Studio interface.
+
 Usually, writing code is a bit quicker, but using the interface can be very useful and complimentary to use of code.
@@
 
-> a package is a like a book, a library is like a library; you use library() to check a package out of the library
+> A package is a like a book, a library is like a library; you use library() to check a package out of the library
 > - Hadley Wickham, Chief Scientist, R Studio
@@
 
-Once you have loaded the package in your session, you can run the functions that are contained within that package. To find a list of all those functions, you can run this in the R Studio console:
+Once you have loaded the package in your session, you can run the functions that are contained within that package.
+
+To find a list of all those functions, you can run this in the R Studio console:
 
@@
 
+Vignettes are long-form documentation (and tutorials) that can provide a helpful introduction to a package.
+
+Run `vignette()` in order to view *all* of the vignettes available for a package:
+
 ```{r, eval = FALSE}
 vignette(package = "dplyr")
+```
+
+Then, you can load a specific vignette.
+
+```{r, eval = FALSE}
 vignette("dplyr", package = "dplyr")
@@
 
+These are also available through CRAN (i.e., https://cran.r-project.org/web/packages/dplyr/index.html)
 ---
@@
 ```{r, eval = FALSE}
-??dplyr::filter
+?dplyr::filter
 ```
@@
 dat <- # example data frame
-  data.frame(stringsAsFactors=FALSE,
-             letter = c("A", "A", "A", "B", "B"),
+  tibble(letter = c("A", "A", "A", "B", "B"),
          number = c(1L, 2L, 3L, 4L, 5L))
 
-dat
-
-filter(dat, letter == "A") # using dplyr::filter
+filter(dat, letter == "A")
 ```
@@
 
-One important consideration is whether to save your workspace when you close R Studio.
-By default, R Studio saves all of the objects in your environment.
-This means that any data that you have loaded--or new data or objects that you have created, such as by merging two data sets together or creating a plot--will, by default, still exist when you open R Studio next.
-In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code.
-This means that when you share your code, or re-run it from the start, it may not work.
-An easy way to change this is to tell R Studio to start from scratch (in terms of your workspace) each time you open it.
-You can do that by changing the dropdown menu pointed out in the image below to "Never".
+By default, R Studio saves all of the objects in your environment. In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code.
 
-![optional caption text](images/save-workspace-reminder.jpg)
----
-
-class: inverse, center, middle
-# Processing data
+```{r, out.width = "425px", echo= FALSE, fig.align = "center"}
+include_graphics("img/save-workspace-reminder.jpg")
+```
 
@@
 
-# The Tidyverse
+# The tidyverse
 
-The Tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data.
+The tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data.
 Tidy data has a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.
@@
 
-**For more information on tidy data, check out [Hadley Wickhams's Tidy Data paper](http://vita.had.co.nz/papers/tidy-data.html).**
----
-
-# Loading Data from Various Sources
-
-### Saving a File from the Web
-
-You'll need to copy this URL:
-
-`https://goo.gl/bUeMhV`
-
-Here's what it resolves to (it's a CSV file):
-
-`https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`
-
-```{r, eval = FALSE}
-student_responses_url <-
-  "https://goo.gl/bUeMhV"
-
-student_responses_file_name <-
-  paste0(getwd(), "./data/student-responses-data.csv")
-
-download.file(
-  url = student_responses_url,
-  destfile = student_responses_file_name)
-```
-
-It may take a few seconds to download as it's around 20 MB.
-
-The process above involves many core data science ideas and ideas from programming/coding. We will walk through them step-by-step.
-
----
-
-# Step 1
-
-1. The *character string* `"https://goo.gl/wPmujv"` is being saved to an *object* called `student_responses_url`.
-
-```{r}
-student_responses_url <-
-  "https://goo.gl/bUeMhV"
-```
-
----
-
-# Step 2
-
-2.
-We concatenate your working directory file path to the desired file name for the CSV using a *function* called `paste0`.
-This is stored in another *object* called `student_reponses_file_name`.
-This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.
-
-```{r}
-student_responses_file_name <-
-  paste0(getwd(), "./data/student-responses-data.csv")
-```
-
----
-
-# Step 3
-
-3. The `student_responses_url` *object* is passed to the `url` argument of the *function* called `download.file()` along with `student_responses_file_name`, which is passed to the `destfile` argument.
-
-In short, the `download.file()` function needs to know
-- where the file is coming from (which you tell it through the `url`) argument and
-- where the file will be saved (which you tell it through the `destfile` argument).
-
-```{r, eval = FALSE}
-download.file(
-  url = student_responses_url,
-  destfile = student_responses_file_name)
-```
-
-Understanding how R is working in these terms can be helpful for troubleshooting and reaching out for help.
-It also helps you to use functions that you have never used before because you are familiar with how some functions work.
-
----
-
-# Loading a CSV File
-
-Okay, we're ready to go. The easiest way to read a CSV file is with the function `read_csv()` from the package `readr`, which is contained within the Tidyverse.
-
----
-
-# Loading the file
-
-Let's load the tidyverse library:
-
-```{r, message = F}
-library(tidyverse) # so tidyverse packages can be used for analysis
-```
-
-After loading the tidyverse packages, we can now load a file. We are going to call the data `student_responses`:
-
-```{r, eval = TRUE}
-student_responses <-
-  read_csv("./data/student-responses-data.csv")
-```
-
-Since we loaded the data, we now want to look at it. We can type its name in the function `glimpse()` to print some information on the dataset (this code is not run here).
-
-```{r, eval = FALSE}
-glimpse(student_responses)
-```
-
----
-
-# Loading Excel files
-
-```{r, eval = FALSE}
-install.packages("readxl")
-```
-
-```{r}
-library(readxl)
-```
-
-We can then use the function `read_excel()` in the same way as `read_csv()`, where "path/to/file.xlsx" is where an Excel file you want to load is located (note that this code is not run here):
-
-```{r, eval = FALSE}
-my_data <-
-  read_excel("path/to/file.xlsx")
-```
-
----
-
-# Loading SAV files
-
-```{r, eval = FALSE}
-install.packages("haven")
-```
-
-```{r, eval = FALSE}
-library(haven)
-my_data <-
-  read_sav("path/to/file.xlsx")
-```
-
----
-
-# Saving Files
-
-Using our data frame `student_responses`, we can save it as a CSV (for example) with the following function.
-The first argument, `student_reponses`, is the name of the object that you want to save.
-The second argument, `student-responses.csv`, what you want to call the saved dataset.
-
-```{r, eval = FALSE}
-write_csv(student_responses, "student-responses.csv")
-```
-
-That will save a CSV file entitled `student-responses.csv` in the working directory.
-If you want to save it to another directory, simply add the file path to the file, i.e.
-`path/to/student-responses.csv`.
-To save a file for SPSS, load the haven package and use `write_sav()`.
-There is not a function to save an Excel file, but you can save as a CSV and directly load it in Excel.
-
----
-class: inverse, center, middle
-# Processing Data
----
-
-# The pipe: %>%
-
-We're also going to introduce a powerful, unusual *operator* in R, the pipe.
-The pipe is this symbol: `%>%`.
-It lets you *compose* functions.
-It does this by passing the output of one function to the next.
-A handy shortcut for writing out `%>%` is Command + Shift + M.
-
-Here's an example.
-Let's say that we want to select a few variables from the `student_responses` dataset and save those variables into a new object, `student_mot_vars`.
-Here's how we would do that using `dplyr::select()`.
-
-```{r}
-student_mot_vars <- # save object student_mot_vars by...
-  student_responses %>% # using dataframe student_responses
-  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
-```
-
----
-
-# Saving the results
-
-Note that we saved the output from the `select()` function to `student_mot_vars` but we could also save it back to `student_responses`, which would simply overwrite the original data frame (the following code is not run here):
-
-```{r, eval = FALSE}
-student_responses <- # save object student_responses by...
-  student_responses %>% # using dataframe student_responses
-  select(student_responses, SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
-```
-
----
-
-# Renaming
-
-We can also rename the variables at the same time we select them.
-I put these on separate lines so I could add the comment, but you could do this all in the same line, too.
-It does not make a difference in terms of how `select()` will work.
-
-```{r}
-student_mot_vars <- # save object student_mot_vars by...
-  student_responses %>% # using dataframe student_responses
-  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
-         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
-         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
-         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
-         student_instrumental_motivation = INSTSCIE # selecting variable INSTSCIE and renaming to student_instrumental_motivation
-  )
-```
-
----
-
-# Processing data (part 1 of 4)
-
-* `select()`
-* `rename()`
-* `filter()`
-* `mutate()`
-
----
-
-# Processing data (part 2 of 4)
-
-* `group_by()`
-* `count()`
-* `summarize()`
-
----
-
-# Processing data (part 3 of 4)
-
-* `left_join()`
-* `right_join()`
-* `semi_join()`
-* `anti_join()`
-
----
-
-# Processing data (part 4 of 4)
-
-* `gather()`
-* `spread()`
-
 ---
@@
 
-# Modeling data
-
----
-
-# Linear models
-
----
-
-# Multi-level models
-
----
-
-# Latent variable models
-
----
-
-class: inverse, center, middle
-# Creating reproducible reports
-
----
-
-# R Markdown
-
----
-
-# APA Tables
-
----
-
-# Papaja
-
----
-
-# Redoc
-
----
-
-class: inverse, center, middle
-
-# Computer programming basics
-
----
-
-# Objects and variables
-
-```{r}
-my_object <- 1
-my_other_object <- c(1:10)
-```
-
-Reading in a CSV and assigning it to an object
-
-```{r}
-my_df <- read_csv("example-data.csv")
-my_df
-```
-
+# Computer science foundations
 ---
@@
 ```{r}
-library(tibble)
 tibble(var1 = vec1,
@@
 ```
+
+---
 
@@
 ```
+
+---
+
+# Conditional logic
 
@@
 ---
+# Objects, functions, directories, and data structures
 
-# Functions
+Here's a short URL: `https://goo.gl/bUeMhV`
 
-```{r}
-mean(c(1, 2, 3))
-```
+This is what it resolves to (it's a CSV file): `https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`
 
-```{r}
-mean(c(1, 2, 3, NA))
-```
+This chunk of code will save that file to a data directory in our working environment.
 
 ```{r, eval = FALSE}
-?mean
-```
+student_responses_url <-
+  "https://goo.gl/bUeMhV"
 
-```{r}
-mean(c(1, 2, 3, NA), na.rm = TRUE)
+student_responses_file_name <-
+  paste0(getwd(), "/data/student-responses-data.csv")
+
+download.file(
+  url = student_responses_url,
+  destfile = student_responses_file_name)
 ```
@@
 
-# Functions
+# Step 1
+
+1. The *character string* `"https://goo.gl/wPmujv"` is being saved to an *object* called `student_responses_url`.
 
 ```{r, eval = FALSE}
-library(tidyverse)
-read_directory <- function(directory) {
-  map_df(list.files(directory), read_csv)
-}
-
-d <- read_directory("data")
+student_responses_url <-
+  "https://goo.gl/bUeMhV"
 ```
@@
 
-# Iteration
+# Step 2
+
+2.
+We concatenate the working directory file path to the desired file name for the CSV using a *function* called `paste0`.
+This is stored in another *object* called `student_reponses_file_name`.
+This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.
+
+```{r, eval = FALSE}
+student_responses_file_name <-
+  str_c(getwd(), "./data/student-responses-data.csv")
+```
+
+---
+
+# Step 3
+
+3. The `student_responses_url` *object* is passed to the `url` argument of the *function* called `download.file()` along with `student_responses_file_name`, which is passed to the `destfile` argument.
+
+In short, the `download.file()` function needs to know
+- where the file is coming from (which you tell it through the `url`) argument and
+- where the file will be saved (which you tell it through the `destfile` argument).
+
+```{r, eval = FALSE}
+download.file(
+  url = student_responses_url,
+  destfile = student_responses_file_name)
+```
+
+---
+class: inverse, center, middle
+# Processing Data
+---
+
+# The pipe: %>%
+
+First, let's load the data we downloaded in the last step.
+
+```{r, message = FALSE}
+student_responses <- read_csv("data/student-responses-data.csv")
+```
+
+We're also going to introduce a powerful, unusual *operator* in R, the pipe.
+The pipe is this symbol: `%>%`.
+It lets you *compose* functions.
+It does this by passing the output of one function to the next.
+
+Select reduces the number of *columns* of a dataset.
+
+```{r, eval = FALSE}
+student_mot_vars <- # save object student_mot_vars by...
+  student_responses %>% # using dataframe student_responses
+  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
+
+student_mot_vars
+```
+
+Also, check out the helper functions: `contains()`, `starts_with()`, and `ends_with()`, i.e.:
+
+```{r, eval = FALSE}
+student_responses %>% 
+  select(contains("sci"))
+```
+
+---
+
+# Saving the results
+
+Note that we saved the output from the `select()` function to `student_mot_vars` but we could also save it back to `student_responses`, which would simply overwrite the original data frame (the following code is not run here):
+
+```{r, eval = FALSE}
+student_responses <- # save object student_responses by...
+  student_responses %>% # using dataframe student_responses
+  select(student_responses, SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
+```
+
+---
+
+# Renaming
+
+We can also rename the variables at the same time we select them.
+I put these on separate lines so I could add the comment, but you could do this all in the same line, too.
+It does not make a difference in terms of how `select()` will work.
+We are using `head()` to only print a few rows.
 
 ```{r}
-random_numbers <- rnorm(n = 10, mean = 0)
-random_numbers
-mean(random_numbers)
-sd(random_numbers)
+student_responses %>% # using dataframe student_responses
+  select(student_efficacy = SCIEEFF,
+         student_joy = JOYSCIE, 
+         student_broad_interest = INTBRSCI,
+         student_epistemic_beliefs = EPIST,
+         student_instrumental_motivation = INSTSCIE,
+  ) %>% 
+  head(3)
+```
+
+---
+
+# Creating a new variable
+
+`mutate()` takes instructions to create a new variable.
+
+What goes *before* the equals sign is the name of the new variable.
+
+What goes after are the instructions.
+
+```{r}
+student_responses %>% 
+  select(student_efficacy = SCIEEFF, 
+         student_joy = JOYSCIE, 
+         student_broad_interest = INTBRSCI, 
+         student_epistemic_beliefs = EPIST,
+         student_instrumental_motivation = INSTSCIE) %>% 
+  mutate(student_joy_interest = (student_joy + student_broad_interest) / 2) %>% 
+  head(3)
+```
+
+---
+
+# Filtering the data set
+
+Filter takes *logical statements* (statements that can evalute to true or false) to select a number of *rows* from a dataset.
+
+```{r}
+student_responses %>% 
+  select(student_efficacy = SCIEEFF, 
+         student_joy = JOYSCIE, 
+         student_broad_interest = INTBRSCI, 
+         student_epistemic_beliefs = EPIST,
+         student_instrumental_motivation = INSTSCIE) %>% 
+  mutate(student_joy_interest = (student_joy + student_broad_interest) / 2) %>% 
+  filter(student_joy_interest > mean(student_joy_interest, na.rm = TRUE)) %>% 
+  head(3)
+```
+
+---
+
+# Grouping and summarizing
+
+It is often useful to aggregate a data set.
+
+The `group_by()` and `summarize()` functions are especially helpful for this purpose.
+
+Let's find the mean and standard deviation (and counts) of student broad interest by teacher.
+
+```{r}
+student_responses %>% 
+  select(student_broad_interest = INTBRSCI, teacher_id = SCHID) %>% 
+  group_by(teacher_id) %>% 
+  summarize(student_broad_interest_mean = mean(student_broad_interest, na.rm = TRUE),
+            student_broad_interest_sd = sd(student_broad_interest, na.rm = TRUE),
+            n = n()) %>% 
+  head(3)
+```
+
+---
+
+# Arranging
+
+Finally, let's arrange the table by the number of students in each class.
+
+`arrange()` sorts in order from the smallest to largest values.
+
+`desc()` tells `arrange()` to sort in descending order.
+
+```{r}
+student_responses %>% 
+  select(student_broad_interest = INTBRSCI, teacher_id = SCHID) %>% 
+  group_by(teacher_id) %>% 
+  summarize(student_broad_interest_mean = mean(student_broad_interest, na.rm = TRUE),
+            student_broad_interest_sd = sd(student_broad_interest, na.rm = TRUE),
+            n = n()) %>% 
+  arrange(desc(n)) %>% 
+  head(3)
+```
+
+---
+
+# Counting
+
+Sometimes, we simply want to count the number of observations associated with each group.
+
+```{r, eval = FALSE}
+student_responses %>% 
+  select(teacher_id = SCHID) %>% 
+  count(teacher_id)
+```
+
+We could then use this as the basis of *other* summary statistics.
+
+```{r}
+student_responses %>% 
+  select(teacher_id = SCHID) %>% 
+  count(teacher_id) %>% 
+  summarize(n_mean = mean(n),
+            n_sd = sd(n))
+```
+
+On average, there are around 32 (*SD* = 8.61) students in each teachers' class.
+
+---
+
+class: inverse, center, middle
+# Demo doc (part 1): Try it out!
+
+---
+
+class: inverse, center, middle
+
+# Loading data
+
+---
+
+# Loading a CSV File
+
+Okay, we're ready to go. The easiest way to read a CSV file is with the function `read_csv()` from the package `readr`, which is contained within the Tidyverse.
+
+Again, we'll load (or have loaded, already) the tidyverse library.
+
+```{r, message = F}
+library(tidyverse) # so tidyverse packages can be used for analysis
+```
+
+Just as a recap of a line that we ran earlier.
+
+```{r, eval = TRUE, message = FALSE}
+student_responses <- read_csv("data/student-responses-data.csv")
+```
+
+Since we loaded the data, we now want to look at it. We can type its name in the function `glimpse()` to print some information on the dataset (this code is not run here).
+
+```{r, eval = FALSE}
+glimpse(student_responses)
+```
+
+You can also print the name of the object.
+
+---
+
+# Loading Excel and SAV files
+
+## Loading Excel files
+
+```{r, eval = FALSE}
+install.packages("readxl")
 ```
@@
 ```{r}
-library(tidyverse)
-1:10 %>%
-  map(~ rnorm(n = 10, mean = .x)) %>%
-  map_dbl(mean)
+library(readxl)
 ```
+
+```{r, eval = FALSE}
+my_data <-
+  read_excel("path/to/file.xlsx")
+```
+
+## Loading SAV files
+
+```{r, eval = FALSE}
+install.packages("haven")
+```
+
+```{r, eval = FALSE}
+library(haven)
+my_data <-
+  read_sav("path/to/file.xlsx")
+```
+
+---
+
+# Saving (Writing) Files
+
+Using our data frame `student_responses`, we can save it as a CSV (for example) with the following function.
+The first argument, `student_reponses`, is the name of the object that you want to save.
+The second argument, `student-responses.csv`, what you want to call the saved dataset.
+
+```{r, eval = FALSE}
+write_csv(student_responses, "student-responses.csv")
+```
+
+That will save a CSV file entitled `student-responses.csv` in the working directory.
+If you want to save it to another directory, simply add the file path to the file, i.e.
+`path/to/student-responses.csv`.
+To save a file for SPSS, load the haven package and use `write_sav()`.
+There is not a function to save an Excel file, but you can save as a CSV and directly load it in Excel.
+
+---
+
+class: inverse, center, middle
+
+# Visualizing data
+
+---
+
+*ggplot2* is a tidyverse package for creating visualizations (or figures).
+
+It is very flexible ... but can also be a bit challenging to use, at first.
+
+Let's work with a smaller data set, for now.
+
+```{r}
+student_mot_vars <- # save object student_mot_vars by...
+  student_responses %>% # using dataframe student_responses
+  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
+         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
+         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
+         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
+         student_instrumental_motivation = INSTSCIE,  # selecting variable INSTSCIE and renaming to student_instrumental_motivation
+         teacher_id = SCHID
+  )
+```
+
+---
+
+# Scatter plot
+
+```{r, warning = FALSE}
+ggplot(student_mot_vars,
+       aes(x = student_efficacy, y = student_broad_interest)) +
+  geom_point()
+```
+
+---
+
+# Scatter plot with a regression line
+
+```{r, warning = FALSE}
+ggplot(student_mot_vars,
+       aes(x = student_efficacy, y = student_broad_interest)) +
+  geom_point() +
+  geom_smooth(method = "lm")
+```
+
+---
+
+# Scatter plot with a regression line and some cleaning up
+
+```{r, warning = FALSE}
+ggplot(student_mot_vars,
+       aes(x = student_efficacy, y = student_broad_interest)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  theme_bw() +
+  xlab("Student Efficacy") +
+  ylab("Student Broad Interest")
+```
+
+---
+
+# Scatter plot with a regression line and some cleaning up
+
+```{r, warning = FALSE}
+ggplot(student_mot_vars,
+       aes(x = student_efficacy, y = student_broad_interest)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  theme_bw() +
+  xlab("Student Efficacy") +
+  ylab("Student Broad Interest")
+```
+
+---
+
+class: inverse, center, middle
+# Demo doc (part 2)
+
+---
+
+class: inverse, center, middle
+
+# Modeling data
+
+---
+
+# Linear models
+
+---
+
+# Multi-level models
+
+---
+
+# Latent variable models
+
+---
+
+class: inverse, center, middle
+# Demo doc (part 3)
+
+---
+
+class: inverse, center, middle
+# Questions, other
+
+---
+
+# R Markdown
+
+---
+
+# APA Tables
+
+---
+
+# Papaja
+
+---
+
+# Redoc
 

### 2019-05-16 demo-doc.Rmd
update

@@
 ```
+
+Let's save the results back to `gss_cat` and make sure that it worked.
+
+```{r}
+gss_cat <- gss_cat %>% 
+  mutate(potential_senior = if_else(age >= 60, TRUE, FALSE))
+```
+
+**Try it out!**
+
+Can you create a new variable based upon either the *year* or *tvhours* variable using `if_else()`?
+
+```{r}
+gss_cat %>% 
+  mutate()
+```
+
+Then, save the results back to `gss_cat` (check out the code two chunks above for an example!).
+
+```{r}
+gss_cat %>% 
+  mutate()
+```
+
+# Demo doc (part 2): Try it out!
+
+`geom_jitter()` is exactly like geom_point, but with a small amount of x- and y-axis random noise added.
+
+```{r}
+ggplot(gss_cat, aes(x = age, y = tvhours)) +
+  geom_jitter()
+```
+
+**Try it out!**
+
+Can you add a line representing the linear relationship between the two varibles? Add another layer using a `geom_` function.
+
+```{r}
+# hint: check out how a line was added in the presentation?
+
+ggplot(gss_cat, aes(x = age, y = tvhours)) +
+  geom_jitter() +
+  geom_
+```
+
+Let's try a different theme.
+
+```{r}
+ggplot(gss_cat, aes(x = age, y = tvhours)) +
+  geom_jitter() +
+  theme_bw()
+```
+
+**Try it out!**
+
+Can you add a different theme? Start typing `theme_` and see what shows up! Be sure to run the code to see the results!
+
+```{r}
+ggplot(gss_cat, aes(x = age, y = tvhours)) +
+  geom_jitter() +
+  theme_
+```
+
+Does the relationship look to be different by year?
+
+```{r}
+ggplot(gss_cat, aes(x = age, y = tvhours)) +
+  geom_jitter() +
+  facet_wrap(~year)
+```
+
+**Try it out!**
+
+Can you add lines of best fit to the above plots?
+
+```{r}
+ggplot(gss_cat, aes(x = age, y = tvhours)) +
+  geom_jitter() +
+  geom_
+  facet_wrap(~year)
+```
+
+# Demo doc (part 3): Try it out!
+
+Let's run a regression.
+
+```{r}
+m1 <- lm(tvhours ~ age, summary = gss_ct)
+summary(m1)
+```
+
+**Try it out!**
+
+Can you add another variable to the model? If you need to check what variables are in `gss_cat`, print that object to the console.
+
+```{r}
+m2 <- lm(tvhours ~ age + , summary = gss_ct)
+summary(m1)
+```

### 2019-05-16 msu-workshop-2019.Rmd
update

@@
 subtitle: "An introduction to R for data science in education"
-author: "Joshua Rosenberg"
+author: "Joshua Rosenberg and Emily Bovee"
 date: "`r Sys.Date()`"
@@
 1. Learning and doing more with R
-1. Questions and sundry tools (time permitting)
+1. Questions
 
@@
 * It is cross-platform, open-source, and freely-available
-* It can h
+
 ---
@@
 > - Hadley Wickham, Chief Scientist, R Studio
-
----
-
-# Running Functions from Packages
-
-Once you have loaded the package in your session, you can run the functions that are contained within that package.
-
-To find a list of all those functions, you can run this in the R Studio console:
-
-```{r, eval = FALSE}
-help(package = dplyr)
-```
-
-The documentation should tell you what the function does, what arguments (i.e., details) needed for it to successfully run, examples, and what the output should look like.
 
@@
 
-We can also rename the variables at the same time we select them.
-I put these on separate lines so I could add the comment, but you could do this all in the same line, too.
-It does not make a difference in terms of how `select()` will work.
-We are using `head()` to only print a few rows.
+We can also rename the variables at the same time we select them. Let's save the results to `student_mot_vars`.
 
 ```{r}
-student_responses %>% # using dataframe student_responses
+student_mot_vars <- student_responses %>% 
   select(student_efficacy = SCIEEFF,
@@
          student_instrumental_motivation = INSTSCIE,
-  ) %>% 
+  )
+
+student_mot_vars %>% 
   head(3)
@@
 
+*Note. Check out [R Studio Cloud](https://rstudio.cloud/) if you're having any installation trouble!*
+
 ---
@@
 
+# The grammar of graphics
+
 *ggplot2* is a tidyverse package for creating visualizations (or figures).
-
-It is very flexible ... but can also be a bit challenging to use, at first.
 
@@
 
-```{r, warning = FALSE}
+Notice the three parts of all `ggplot2` graphs: a) the data (`student_mot_vars`), b) an `aes()`thetic mapping, and c) a `geom`:
+
+```{r, warning = FALSE, out.width = "450px", echo = FALSE, fig.align = "center"}
 ggplot(student_mot_vars,
@@
 
-```{r, warning = FALSE}
+```{r, warning = FALSE, out.width = "450px", fig.align = "center"}
 ggplot(student_mot_vars,
@@
 
-```{r, warning = FALSE}
-ggplot(student_mot_vars,
-       aes(x = student_efficacy, y = student_broad_interest)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  theme_bw() +
-  xlab("Student Efficacy") +
-  ylab("Student Broad Interest")
-```
-
----
-
-# Scatter plot with a regression line and some cleaning up
-
-```{r, warning = FALSE}
+```{r, warning = FALSE, out.width = "450px", fig.align = "center"}
 ggplot(student_mot_vars,
@@
 
+The `lm()` function is a very helpful general purpose function for linear models. We'll use the `student_mot_vars` data frame we created earlier.
+
+```{r}
+lm(student_epistemic_beliefs ~ student_broad_interest, data = student_mot_vars)
+```
+
+---
+
+# Linear models
+
+Let's save the results back to an object, `m1`.
+
+```{r}
+m1 <- lm(student_epistemic_beliefs ~ student_broad_interest, data = student_mot_vars)
+```
+
+---
+
+# Linear models
+
+We can then run a summary on the output.
+
+```{r}
+summary(m1)
+```
+
+---
+
+# Linear models
+
+You can then build up a more complex model, i.e.:
+
+```{r, eval = FALSE}
+m2 <- lm(student_epistemic_beliefs ~ student_broad_interest + student_joy + student_broad_interest:student_joy, data = student_mot_vars)
+```
+
+You can then run `summary()` to view the results.
+
 ---
@@
 
----
-
-# Latent variable models
-
+```{r, eval = TRUE}
+library(lme4)
+m2_mlm <- lmer(student_epistemic_beliefs ~ student_broad_interest + student_joy + student_broad_interest:student_joy + (1|teacher_id), data = student_mot_vars)
+m2_mlm
+```
 ---
@@
 class: inverse, center, middle
-# Questions, other
-
----
-
-# R Markdown
-
----
-
-# APA Tables
-
----
-
-# Papaja
-
----
-
-# Redoc
-
----
-
 # Learning and doing more with R
@@
 
-# Following up
+# Resources
+
+  - [Doing data science with R](https://r4ds.had.co.nz/) by Wickham and
+    Grolemund (2017)
+  - [Big magic with R: Creating learning beyond
+    fear](https://speakerdeck.com/apreshill/big-magic-with-r-creative-learning-beyond-fear)
+    by Hill (2017)
+  - [Data science in education](https://github.com/data-edu/data-science-in-education)
+
+---
+
+# Courses
+
+  - [\#r4ds](https://medium.com/@kierisi/r4ds-the-next-iteration-d51e0a1b0b82)
+    (see a talk at rstudio::conf()
+    [here](https://resources.rstudio.com/rstudio-conf-2019/r4ds-online-learning-community-improvements-to-self-taught-data-science-and-the-critical-need-for-diversity-equity-and-inclusion-in-data-science-education)
+    by Maegan (2019))
+  - [Data science for social scientists](http://datascience.tntlab.org/) by
+    Landers (2019)
+  - [University of Oregon Data Science Specialization for the College of
+    Education](https://github.com/uo-datasci-specialization) by Anderson (2019)
+
+---
+
+class: inverse, center, middle
+# Questions?
+
+The repository for this workshop is [here](https://github.com/jrosen48/MSU-workshop-2019).

### 2019-05-19 demo-doc.Rmd
update typos/edits based on presentation

@@
 gss_cat %>% 
+  select(year, tvhours)
+
+gss_cat %>% 
   select(year, marital, tvhours)
@@
 gss_cat %>% 
-  filter(age > 30)
+  filter(age > 60)
 ```
@@
 gss_cat %>% 
-  mutate(potential_senior = if_else(age >= 60, TRUE, FALSE))
+  mutate(potential_senior = if_else(age >= 63.5, 1, 0),
+         super_senior = if_else(potential_senior == 1 & tvhours < 2, 1, 0)) %>% 
+  count(potential_senior, super_senior)
 ```
@@
 gss_cat <- gss_cat %>% 
-  mutate(potential_senior = if_else(age >= 60, TRUE, FALSE))
+  mutate(potential_senior = ifelse(age >= 63.5, TRUE, FALSE))
+```
+
+Adding another variable for folks who are middle-aged.
+
+```{r}
+gss_cat <- gss_cat %>% 
+  mutate(middle_aged = ifelse(age >= 30 & age <= 40, "yes", "no"))
 ```
@@
 ```{r}
-m1 <- lm(tvhours ~ age, summary = gss_ct)
+m1 <- lm(tvhours ~ age, data = gss_cat)
 summary(m1)
@@
 ```{r}
-m2 <- lm(tvhours ~ age + , summary = gss_ct)
+m2 <- lm(tvhours ~ age + , data = gss_cat)
 summary(m1)

### 2019-05-20 msu-workshop-2019.Rmd
update

@@
 author: "Joshua Rosenberg and Emily Bovee"
-date: "`r Sys.Date()`"
+date: "2018-05-16"
 output:
@@
 
-"Packages" are shareable collections of R code that provide functions (i.e., a command to perform a specific task), data and documentation,.
+"Packages" are shareable collections of R code that provide functions (i.e., a command to perform a specific task), data, and documentation.
 Packages increase the functionality of R by improving and expanding on base R (basic R functions).
@@
 d1 %>% 
-  mutate(var2 = ifelse(vec1 >= .5, "greater", "lesser"))
+  mutate(var2 = ifelse(var1 >= .5, "greater", "lesser"))
 ```
@@
     var1 >= .75 ~ "a lot",
-    var1 >= .5 ~ "greater",
-    var1 < .5 ~ "lesser",
-    var1 <= .25 ~ "very little" 
+    var1 >= .5 & var1 < .75 ~ "greater",
+    var1 >= .25 & var1 < .5  ~ "lesser",
+    var1 < .25 ~ "very little" 
   ))
@@
 2.
-We concatenate the working directory file path to the desired file name for the CSV using a *function* called `paste0`.
+We concatenate the working directory file path to the desired file name for the CSV using a *function* called `str_c`.
 This is stored in another *object* called `student_reponses_file_name`.
@@
 
+*The doc is [here](https://github.com/jrosen48/MSU-workshop-2019/blob/master/demo-doc.Rmd) in the repository*
+
 ---
@@
 
-```{r, warning = FALSE, out.width = "450px", fig.align = "center"}
+```{r, warning = FALSE, out.width = "350px", fig.align = "center"}
 ggplot(student_mot_vars,
@@
 ```
+---
+# Cleaning up
 
----
-
-# Scatter plot with a regression line and some cleaning up
-
-```{r, warning = FALSE, out.width = "450px", fig.align = "center"}
+```{r, warning = FALSE, out.width = "350px", fig.align = "center"}
 ggplot(student_mot_vars,
@@
 ```
-
 ---
@@
 # Demo doc (part 2)
+
+*The doc is [here](https://github.com/jrosen48/MSU-workshop-2019/blob/master/demo-doc.Rmd) in the repository*
 
@@
 
+*The doc is [here](https://github.com/jrosen48/MSU-workshop-2019/blob/master/demo-doc.Rmd) in the repository*
 ---
@@
 The repository for this workshop is [here](https://github.com/jrosen48/MSU-workshop-2019).
+
+---
+
+# Contact
+
+- [Joshua Rosenberg](mailto:jmrosenberg@utk.edu)
+- [Emily Bovee](mailto:emily.a.bovee@gmail.com)

### 2019-05-20 msu-workshop-2019.Rmd
update

@@
   - [Data science in education](https://github.com/data-edu/data-science-in-education)
+  - [R Studio Community](https://community.rstudio.com/)
 
@@
 
+# Useful packages
+
+With one exception, can install via `install.packages(<name-of-package)`, i.e., `install.packages("tidylog").
+
+- tidyverse: tools for data manipulation and processing
+- tidylog: companion to the tidyverse; gives you feedback
+- haven: import and export files for other statistical software
+- apaTables: create APA-style tables in MS Word format
+- devtools: to download packages only available on GitHub
+- psych: general psychological science-related analysis tools
+- skimr: quickly calculate summary statistics for a data frame
+- lavaan: structural equation modeling
+- tipyLPA: latent profile analysis
+- lme4: multi-level modeling (HLM)
+- sjstats: convenient statistical functions
+- poLCA: latent class analysis
+- rmarkdown: reports (PDF, Word, HTML, etc.)
+- papaja: APA-style paper generation (must install from [GitHub](https://github.com/crsh/papaja))
+- blogdown create a website/blog
+- xaringan: create presentations
+- MplusAutomation: Run Mplus via R
+
+---
+
+# People (and a few hashtags) to follow (on Twitter)
+
+- [Jesse Mostipak](https://twitter.com/kierisi)
+- [Alison Hill](https://twitter.com/apreshill)
+- [Hadley Wickham](https://twitter.com/hadleywickham)
+- [Daniel Anderson](https://twitter.com/datalorax_)
+- [Mara Averick](https://twitter.com/dataandme)
+- [Thomas Mock](https://twitter.com/thomas_mock)
+- [Angela Li](https://twitter.com/CivicAngela)
+- [R4DS community](https://twitter.com/R4DScommunity)
+- [#tidytuesday](https://twitter.com/hashtag/TidyTuesday?src=hash)
+- [#rstats](https://twitter.com/hashtag/rstats?src=hash)
+- [David Robinson](https://twitter.com/drob)
+- [Julia Silge](https://twitter.com/juliasilge)
+- [Gabriela de Quieroz](https://twitter.com/gdequeiroz)
+- [Joshua Rosenberg](https://twitter.com/jrosenberg6432)
+- [Emily Bovee](https://twitter.com/ebovee09)
+
+---
+
 class: inverse, center, middle

### 2019-05-21 msu-workshop-2019.Rmd

@@
 options(htmltools.dir.version = FALSE)
+
+# rmarkdown::render("msu-workshop-2019.rmd", output_file = 'index.html')
 
@@
 - [Gabriela de Quieroz](https://twitter.com/gdequeiroz)
-- [Joshua Rosenberg](https://twitter.com/jrosenberg6432)
-- [Emily Bovee](https://twitter.com/ebovee09)
+- [Joshua Rosenberg](https://twitter.com/jrosenberg6432) and [Emily Bovee](https://twitter.com/ebovee09) (of course! :)
 

### 2019-09-04 mtsu-workshop-2019.Rmd
update

--- 2019-05-20 msu-workshop-2019.Rmd
@@
 ---
-title: "Won’t you be my neighboR?"
-subtitle: "An introduction to R for data science in education"
-author: "Joshua Rosenberg and Emily Bovee"
-date: "2018-05-16"
+title: "An Introduction to Data Science in Education in Education Using R"
+subtitle: ""
+author: "Joshua Rosenberg"
+date: "2019-09-05"
 output:
@@
 options(htmltools.dir.version = FALSE)
+
+# rmarkdown::render("MTSU-workshop.rmd", output_file = 'index.html')
 
@@
 include_graphics("img/kr-joro.jpg")
-```
-
----
-
-# #whoami2
-
-.pull-left[
-* Dr. Emily Bovee
-* Director of Educational Development and Assessment: Marquette University School of Dentistry
-* Cat mom
-* Primary areas of interest
-  * Understanding student motivation in higher education and professional education
-  * Data science education 
-]
-
-```{r, out.width = "350px", echo = FALSE, fig.align = "center"}
-include_graphics("img/ebv.jpg")
 ```
@@
 
-*The doc is [here](https://github.com/jrosen48/MSU-workshop-2019/blob/master/demo-doc.Rmd) in the repository*
+*The doc is [here](https://github.com/jrosen48/MTSU-workshop/blob/master/demo-doc.Rmd) in the repository*
 
@@
 
-*The doc is [here](https://github.com/jrosen48/MSU-workshop-2019/blob/master/demo-doc.Rmd) in the repository*
+*The doc is [here](https://github.com/jrosen48/MTSU-workshop/blob/master/demo-doc.Rmd) in the repository*
 ---
@@
   - [R Studio Community](https://community.rstudio.com/)
+  - [Data Science in Education Using R](https://github.com/data-edu/data-science-in-education)
 
@@
 - [Gabriela de Quieroz](https://twitter.com/gdequeiroz)
+- [Emily Bovee](https://twitter.com/ebovee09)
 - [Joshua Rosenberg](https://twitter.com/jrosenberg6432)
-- [Emily Bovee](https://twitter.com/ebovee09)
 
@@
 
-The repository for this workshop is [here](https://github.com/jrosen48/MSU-workshop-2019).
+The repository for this workshop is [here](https://github.com/jrosen48/MTSU-workshop).
 
@@
 
+# Acknowlegment
+
+This presentation is based on one given with Dr. Emily Bovee: https://github.com/jrosen48/MSU-workshop-2019
+
+---
 # Contact
@@
 - [Joshua Rosenberg](mailto:jmrosenberg@utk.edu)
-- [Emily Bovee](mailto:emily.a.bovee@gmail.com)
